---
title: "CaseStudy2"
author: "Robert Hazell"
date: "4/6/2019"
output: 
  html_document:
    keep_md: true
---

```{r include=FALSE} 
knitr::opts_chunk$set(comment = NA, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(kableExtra)
library(magrittr)
```

```{r}
setwd("~/Desktop/DataScience@SMU/DoingDataScience/CaseStudy_2")
library(readxl)
emp_data <- read_excel("CaseStudy2-data.xlsx")
```

Exploratory questions
- proportion of people married in each job role
- proportion of males and females in each job role
- frequency of travel by job role
- which job roles hold the highest average worker employment length

High level statistics of number of people married in each role.

```{r}
emp_marital <- emp_data %>%
  group_by(JobRole) %>%
  summarise(Married = sum(MaritalStatus == "Married"),
            Single = sum(MaritalStatus == "Single"),
            Divorced = sum(MaritalStatus == "Divorced"))

# tidy to make grouped barplot
emp_marital_revised <- emp_marital %>% 
  gather(Married, Single, Divorced, 
         key = "Marriage Status", value = "Employees")

# plot
ggplot(emp_marital_revised, aes(JobRole, Employees, fill = `Marriage Status`)) +
  geom_bar(stat = "identity") + 
  coord_flip() + 
  ggtitle("Number of Employees in each Marital Status by Job Role") + 
  xlab("Job Role") + 
  labs(fill = "Marriage Status") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", family = "Helvetica"))
```

Immediately we can see Sales Execs, Research Scientists, and Lab Technicians comprise the majority of single, married, and divorced workers.

Do jobs that require more frequent travel also feature a higher number of divorced employees?  Note that causality cannot be inferred here; in other words, we don't know if they're divorced for work-related reasons or not.  We're just seeking to establish correlation.

Alternatively, which job roles feature employees with higher ```RelationshipSatisfaction```?

Let's answer the first question.

```{r}
emp_data %>% 
  group_by(BusinessTravel) %>%
  summarise(Divorced = sum(MaritalStatus == "Divorced")) %>%
  arrange(desc(Divorced)) %>%
  kable(align = rep('c',2)) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Those who travel frequently fall in the middle in terms of divorce numbers, but again we cannot make causal inferences.

Now we'll like to see which job roles hold the highest average worker employment length

```{r}
emp_data %>%
  group_by(JobRole) %>%
  summarise(`Avg Yrs at Company` = round(mean(YearsAtCompany),2)) %>%
  arrange(desc(`Avg Yrs at Company`)) %>%
  kable(align = rep('c',2)) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Not surprisingly, Managers have worked longer at their current company and sales reps the least.

For each job role, do males or females tend to have longer commutes?  Our initial guess is that distance doesn't significantly differ. Let's see:

```{r}
emp_data %>%
  group_by(JobRole, Gender) %>%
  summarise(`Avg Commute Distance` = mean(DistanceFromHome)) %>%
  ggplot(aes(JobRole, `Avg Commute Distance`, fill = Gender)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  ggtitle("Average Commuting Distance by Job Role and Gender") + 
  ylab("Avg Commuting Distance") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", family = "Helvetica"))+
  scale_fill_manual(values = c("steel blue", "orange"))
```

Average commuting distance seems gender-invariant for the most part.

We'd also like to know which job roles include a higher travel burden.

```{r}
emp_data %>%
  filter(BusinessTravel == "Travel_Frequently") %>%
  group_by(JobRole) %>%
  summarise(Employees = n()) %>%
  ggplot(aes(reorder(JobRole, Employees), Employees)) + 
  geom_bar(stat = "identity", fill = "black") + 
  coord_flip() + 
  ggtitle("Frequent Business Travel by Job Role") +
  xlab("Position") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", family = "Helvetica")) + 
  theme(legend.position = "none")
```

We'd like to compare salaries (assessed by ```MonthlyRate```) across all JobRoles)

```{r}
ggplot(emp_data, aes(JobRole, MonthlyIncome, fill = JobRole)) + 
  geom_boxplot() + 
  coord_flip() + 
  theme(legend.position = "none") + 
  ggtitle("Comparison of Monthly Income by Job Role") + 
  ylab("Monthly Income") + 
  xlab("Position") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", family = "Helvetica"))
```

Interestingly, these two pairs respectively:

* Research Director + Manager
* Sales Executive + Manufacturing Director

both have virtually identical average monthly incomes and distributions, that is the range in monthly income for both pairs are approximately the same.  The first pair are noticeably left-skewed while the second pair is right skewed.

Now, we'd like to examine features and trends relating to attrition.  First, how does job involvement correlate to attrition for each job role?

```{r}
# get a data frame summarizing attrition numbers by job involvement for each job role
emp_data$JobInvolvement %<>% as.character()
emp_data %>%
  group_by(JobRole, JobInvolvement) %>%
  summarise(Total = n()) %>%
  as.data.frame() %>%
  ggplot(aes(JobRole, Total, fill = JobInvolvement)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  ggtitle("Employee Attrition reflecting Job Involvement") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", family = "Helvetica")) +
  xlab("Job Role") + ylab("Attrition") + labs(fill = "Involvement")
```

It seems like those on the extremes (Involvement of 1 or 4), attrition is least.  But for those with involvement falling somewhere in the middle (2 and 3), attrition is highest.  


